---
title: "chapter-2"
author: "Sonya Hua"
date: "October 3, 2017"
output: html_document
---

```{r cars}
library(astsa)

# ------------------------------------------------------------------------------------
# Global warming

data(gtemp)
plot(gtemp,type="o",ylab="Global Temperature Deviations")

summary(fit<-lm(gtemp~time(gtemp))) # regress gtemp on time
abline(fit) # add regression line to the plot


```
```{r}

data(soi); data(rec)
par(mfrow=c(2,1))
plot(soi,ylab="",xlab="",main="Southern Oscillation Index")
plot(rec,ylab="",xlab="",main="Recruitment")

par(mfrow=c(3,1))
acf(soi,48,main="Southern Oscillation Index")
acf(rec,48,main="Recruitment")
ccf(soi,rec,48,main="SOI vs Recruitment",ylab="CCF")

fish=ts.intersect(rec,soiL6=lag(soi,-6),dframe=TRUE)
summary(lm(rec~soiL6, data=fish,na.action=NULL))

```

```{r}


# ------------------------------------------------------------------------------------
# Pollution, temperature and mortality

data(cmort); data(tempr); data(part)
par(mfrow=c(3,1))
plot(cmort,main="Cardiovascular Mortality",xlab="",ylab="")
plot(tempr,main="Temperature",xlab="",ylab="")
plot(part,main="Particulates",xlab="",ylab="")

dev.new() # open a new graphic device for the scatterplot matrix
pairs(cbind(Mortality=cmort,Temperature=tempr,Particulates=part))

temp=tempr-mean(tempr) # center temperature
temp2=temp^2
trend=time(cmort) # time
fit=lm(cmort~trend+temp+temp2+part,na.action=NULL)
summary(fit) # regression results
summary(aov(fit)) # ANOVA table (compare to next line)
summary(aov(lm(cmort~cbind(trend,temp,temp2,part)))) # Table 2.1

num=length(cmort) # sample size
AIC(fit)/num-log(2*pi) # AIC
AIC(fit,k=log(num))/num-log(2*pi) # BIC
(AICc=log(sum(resid(fit)^2)/num)+(num+5)/(num-5-2)) # AICc


```
```{r}

# ------------------------------------------------------------------------------------
# Global warming

data(gtemp)
plot(gtemp,type="o",ylab="Global Temperature Deviations")
summary(fit.q<-lm(gtemp~time(gtemp)+I(time(gtemp)^2)))

par(mfrow=c(2,2))
plot(gtemp,type="o",main="gtemp")
lines((1880:2009),fit.q$fitted.values,col="red")
plot(resid(fit),type="o",main="detrended")
acf(gtemp,48,main="gtemp")
acf(resid(fit),48,main="detrended")

par(mfrow=c(3,1))
plot(gtemp,type="o",main="gtemp")
plot(resid(fit),type="o",main="detrended")
plot(diff(gtemp),type="o",main="first difference")

par(mfrow=c(3,1))
acf(gtemp,48,main="gtemp")
acf(resid(fit),48,main="detrended")
acf(diff(gtemp),48,main="first difference")

```
```{r}


# ------------------------------------------------------------------------------------
# Accidental deaths, USA, 1973-1978; deaths.txt

deaths=read.table("C:/Users/sonya/Documents/~Time Series/deaths.txt",sep="")
deaths=as.matrix(deaths/1000); n.d=length(deaths)

year=seq(1973,by=1/12,length=n.d)
plot(year,deaths,type="o",pch=22,lty=1,pty=2,xlab="year",ylab="monthly accidental deaths (in thousands)")

```
*Obs:* Xt = monthly accidental deaths. 


cos/sine regressors are used for harmonic regression
```{r}

# Harmonic regression
t=seq(1,n.d)  # time
lambda.1=2*pi/12; lambda.2=2*pi/6  # lambdas capture the monthly pattern and bi-monthly pattern
cos.1=cos(lambda.1*t); cos.2=cos(lambda.2*t); sin.1=sin(lambda.1*t); sin.2=sin(lambda.2*t)  # regressors defined
data.deaths=data.frame(deaths,cos.1,cos.2,sin.1,sin.2)  # put all data into a data frame to fit the linear regression model

fit.deaths=lm(deaths~1+cos.1+cos.2+sin.1+sin.2,data=data.deaths)
summary(fit.deaths)

plot(year,deaths,type="p",pch=22,lty=1,pty=2,xlab="year",ylab="monthly accidental deaths (in thousands)")
lines(year,fit.deaths$fitted.values,lty=1,col="red")

```
NOTE: This will not be covered in exam. We put both cosine and sine in model. Overall test is highly significant and harmonic model explains 65% of variation.

Plot shows fitted curve with time series plot. Overall model captures the pattern with the exception of 1973 to 1974 which can be due to sampling variation. 

```{r}

# ------------------------------------------------------------------------------------
# Paleoclimatic glacial varves

data(varve)
par(mfrow=c(2,1))
plot(varve,main="varve",ylab="")
plot(log(varve),main="log(varve)",ylab="" )

```
*Obs*: Notice the changing variability betwteen 300 to 500 Time points. We will apply a log transformation to stabilize variation throughout the ts. Log transformation worked pretty well. 

We need to first take care of trend, seasonality, and changing variability. Get rid of changing variability first. After stabilizing changing variability. Then we can talk about trend and seasonality.


```{r}
# ------------------------------------------------------------------------------------
# El Nino and fish population

data(soi); data(rec)
lag1.plot(soi,12)
lag2.plot(soi,rec,8)

```
*Obs*: 12 is the maximum lag as seen from plots. This is used to determine the autoregressors or variables to be included in the autoregressive model. We notice a clear linear relationship at time lag =0 and 1. We also notice this in the sample ACF which is significant. These curves are fitted. In the upper right corner is the correlation coefficient, measure of linear association. Weak correlation is <0.05. We want the time lag that provides us the lowest correlation. 

In the next matrix is for 2 time series. (Yt, Xt). This is the correlation between 2 time series. The relationship is not linear. 

When doing homework, we'll need to use these functions.
```{r}
# ------------------------------------------------------------------------------------
# Moving average smoother

data(cmort)
ma5=filter(cmort,sides=2,rep(1,5)/5)  # To do moving average, we use filter(). K = 2, monthly
ma53=filter(cmort,sides=2,rep(1,53)/53) # K = 26 (weekly)
plot(cmort,type="p",ylab="mortality")
lines(ma5,col="blue"); lines(ma53,col="red")
```
*Obs*: Red trend is considerably a smoother moving average than then blue trend

```{r}
# ------------------------------------------------------------------------------------
# Polynomial and periodic regression smoothers

data(cmort)
wk=time(cmort)-mean(time(cmort)) # wk is essentially t/52 centered at zero
wk2=wk^2; wk3=wk^3
cs=cos(2*pi*wk); sn=sin(2*pi*wk)
reg1=lm(cmort~wk+wk2+wk3,na.action=NULL)
reg2=lm(cmort~wk+wk2+wk3+cs+sn,na.action=NULL)
plot(cmort,type="p",ylab="mortality")
lines(fitted(reg1),col="red"); lines(fitted(reg2),col="blue")
```

```{r}
# ------------------------------------------------------------------------------------
# Kernel smoother

data(cmort)
plot(cmort,type="p",ylab="mortality")
lines(ksmooth(time(cmort),cmort,"normal",bandwidth=5/52),col="blue")
lines(ksmooth(time(cmort),cmort,"normal",bandwidth=2),col="red")


```
*Obs*: At b=2, we have a larger bandwidth and as a result a smoother line (red). 

All 3 methods produce reasonable stationarity. In this case, non-parametric methods seem to outperform parametric methods. 


