---
title: "ch-1-time-series"
author: "Sonya Hua"
date: "September 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
install.packages("astsa")

```{r fig.width=12}
library(astsa)

# ------------------------------------------------------------------------------------
# Johnson & Johnson quarterly earnings

data(jj)
plot(jj,type="o",ylab="Quarterly Earnings per Share")

```
*Obs: *
There is an upward *trend* in the data. 
There is a *seasonal* quarterly pattern (Q3 is highest (peak), Q4 is lowest (trough)). 
There is *changing variability* as the variability increases in later years
There does not appear to be missing data or structural breaks (when there's a sharp change from one part to another part)
There aren't any outliers

```{r fig.width=12}
# ------------------------------------------------------------------------------------
# Global warming (Yearly data starting from 1880 to 2009)
# Freq = 1 for annual data
# Measures deviation from the mean of temperature between 1951 to 1980

data(gtemp)
plot(gtemp,type="o",ylab="Global Temperature Deviations")

```
*Obs*: 
There is an upward trend between 1910 to 1940 then 1940 to 2010
There does not appear to be any seasonality since it's yearly data.
There does not appear to be an changing variability
There may be a structural break between 1935 to 1965 (2 parts from constant to increase). A level jump then a sharp upward trend from 1965 onwards
No missing data
No outliers
```{r fig.width=14}
# ------------------------------------------------------------------------------------
# New York Stock Exchange
# time intervals are In days from 1991 onwards
# xt is returns, with holidays and weekends removed
data(nyse)
plot(nyse,ylab="NYSE Returns")

```
*Obs*: 
No apparent upward/downward trend (slow changes)
No apparent seasonality (hard to tell)
Outlier occurs ~ Day 950 (Market Crash of Oct, 19 1987)
There's changing volatility (conditional variance) of (peaks followed by troughs)
```{r fig.width=14}
# ------------------------------------------------------------------------------------
# El Nino and fish population
# Monthly data, 2 measurements: Oscillation Index, New Fish Recruitment

data(soi); data(rec)
par(mfrow=c(3,1))
plot(soi,ylab="",xlab="",main="Southern Oscillation Index")
plot(rec,ylab="",xlab="",main="Recruitment")

plot(soi,ylab="",xlab="",main="El Nino and fish population")
lines(rec/100,col="blue")  # Adding blue recruitment line in the plot. Divide by 100 to get the same y-axis scale for super-imposement
legend(1950,-0.7,c("soi","rec/100"),col=c("black","blue"),lty=c(1,1),bty="n") # create a legend to distinguish the lines
# legend(x, y=NULL, legend)
# bty: the type of box to be drawn around the legend. The allowed values are "o" (the default) and "n".
# -0.7 is the vertical axis's starting position
```
*Obs* 
There are 2 time series then a combined time series
For both ocsillation and recruitment, there's seasonal monthly cycles
There seems to be strong dependence between oscillation and recruitment based on the overlapped time series

```{r fig.width=12}
# ------------------------------------------------------------------------------------
# Gaussian white noise, MA, & AR

w=rnorm(500) # 500 random obs from a normal distribution

# Moving average with time lag = 3
v=filter(w,rep(1/3,3),sides=1)  
# Filter uses moving average method by default
# 3 elements with .33 weight
# If sides = 1 the filter coefficients are for past values only; if sides = 2 they are centred around lag 0. In this case the length of the filter should be odd, but if it is even, more of the filter is forward in time than backward.

# Autoregression sequence with order 2 (2 coefficients)
x=filter(rnorm(550),filter=c(1,-.9),method="recursive")[-(1:50)]
# autogressed on 550 vs. 500 because we need previous 2 values in order to estimate x each time. Add 50 extra obs to circumvent this issue
# filter = the coefficients or vector of filter coefficients in reverse time order (as for AR or MA coefficients).
# when method = recursive, autoregression is used

par(mfrow=c(3,1))
plot.ts(w,main="Gaussian white noise")
plot.ts(v,main="Moving average")
plot.ts(x,main="Autoregression")
```
*Observe* 
* For gaussian, most of the values are between -3 and 3. There are also no specific patterns in the series. We can also draw a normal probability plot to check normality
* for moving average, we lose some information in the beginning (unstable) since we need previous values in order to predict the next value. These values will show up as missing for first 2 values. The series is smoothed. Values tend to be less extreme due to averagine. 
* For Autoregression, there is a cylical periodicity. 

Each of these models capture certain dependence in the time series. 

```{r}

# ------------------------------------------------------------------------------------
# Random walk

set.seed(154) #psedorandom # generator
w=rnorm(200); x=cumsum(w) # w= white noise, x = cumulative sum
wd=w+.2; xd=cumsum(wd) #wd = with drift (0.2)
plot.ts(xd,ylim=c(-5,55),ylab="",main="random walk") 
lines(x, col="darkblue") # without drift
lines(.2*(1:200),lty="dashed") # add a reference line

```
*Observe*
* Without drift, we should expect a more horizontal randomness
* With drift, we should expect a trend (increasing in this case)
```{r}
# ------------------------------------------------------------------------------------
# Example: sample ACF of 200 simulated values of iid N(0,1) noise

x=rnorm(200)
par(mfrow=c(2,1))
plot.ts(x,type="o",ylab="",main="Gaussian white noise"); abline(h=0)
acf(x,40,main="")
```




# ------------------------------------------------------------------------------------
# SOI and recruitment

data(soi); data(rec)
par(mfrow=c(3,1))
acf(soi,48,main="Southern Oscillation Index")
acf(rec,48,main="Recruitment")
ccf(soi,rec,48,main="SOI vs Recruitment",ylab="CCF")


# ------------------------------------------------------------------------------------
# Soil surface temperatures

data(soiltemp)
persp(1:64,1:36,soiltemp,phi=30,theta=30,scale=FALSE,expand=4,ticktype="detailed",xlab="rows",ylab="cols",zlab="temperature")

fs=abs(fft(soiltemp-mean(soiltemp)))^2/(64*36)
cs=Re(fft(fs,inverse=TRUE)/sqrt(64*36)) # ACVF
rs=cs/cs[1,1] # ACF
rs2=cbind(rs[1:41,21:2],rs[1:41,1:21])
rs3=rbind(rs2[41:2,],rs2)
par(mar=c(1,2.5,0,0)+.1)
persp(-40:40,-20:20,rs3,phi=30,theta=30,expand=30,scale="FALSE",ticktype="detailed",xlab="row lags",ylab="column lags",zlab="ACF")


