---
title: "Homework-1"
author: "Sonya Hua"
date: "September 25, 2017"
output: rmarkdown::github_document
---


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

# set global display parameters
knitr::opts_chunk$set(fig.width=10, fig.height=10, echo = TRUE, fig.align="center") 
```

### Question 1. 

Create a vector for the following data set and provide mean, std dev:  27, 30, 33, 36, 39, 42, 45, 48, 51, and 5

```{r}
x <- c(27, 30, 33, 36, 39, 42, 45, 48, 51, 5)

print(paste("The standard deviation is: ", sd(x)))
print(paste("The mean of vector x is: ", mean(x)))
```
### Question 2

a).  Create a numeric vector x of length 80 that ranges from −pi to pi
b). Create a vector y1 that is the sine of x (in radians)
c). Create a vector y2 that is the cosine of x (in radians)
d).  Plot y1 vs. x as a series of points joined by lines. On the same graph, add red-colored points for y2 vs x. 

```{r}

# a) 
(x <- seq(-pi,pi,length=80))
```

```{r}
# b)
(y1 <- sin(x))

```

```{r}
# c)
(y2 <- cos(x))
```

```{r}
# d)  Plot y1 vs. x as a series of points joined by lines. 
# On the same graph, add red-colored points for y2 vs x.
par(mfrow=c(1,1)) 
plot(y1, x, type="b")
lines(y2,x, type="p", col="red")

```

### Question 3

a). Simulate a completely random process of length 48 with independent normal values. Repeat this exercise three more times with a new simulation each time. 

b). Draw the four time series in a 2-by-2 layout

```{r}
# a)
set.seed(3856)
ts1 <- rnorm(n=48)

set.seed(2345)
ts2 <- rnorm(n=48)

set.seed(9987)
ts3 <- rnorm(n=48)

set.seed(96387)
ts4 <- rnorm(n=48)
```

```{r}
par(mfrow=c(2,2))
plot.ts(ts1)
plot.ts(ts2)
plot.ts(ts3)
plot.ts(ts4)
```



